---
title: "Association Rules — Market Basket Analysis"
---

Given a pile of receipts, *association rule mining* discovers patterns of the form
*"customers who buy A and B also tend to buy C"*. The textbook algorithm is
[Apriori](https://en.wikipedia.org/wiki/Apriori_algorithm) — fast, deterministic, and the standard starting point.

A rule $A \Rightarrow B$ is characterized by three numbers:

- **Support** — fraction of baskets containing $A \cup B$, i.e. $P(A \cup B)$. High support = the rule covers a meaningful share of customers.
- **Confidence** — given $A$ in the basket, how likely is $B$, i.e. $P(B \mid A)$. High confidence = the rule is reliable.
- **Lift** — how much more often $B$ shows up when $A$ is present compared to its baseline rate, i.e. $\frac{P(B \mid A)}{P(B)}$. Lift > 1 = positive association.

We mine rules at minimum support 0.001 and minimum confidence 0.5, then sort by confidence. Two implementations run side by side: `arules` in R and `mlxtend` in Python.

## Data

For market basket analysis we only need two columns: the basket identifier (`transaction_id`) and the item identifier. We use `article_name`, not `article_id` — so a "sofa" rule isn't fragmented across three SKUs of the same product.

```{r}
#| label: load-data-r
library(dplyr)
library(readr)

raw <- read_delim("data/synthetic/transactions.csv", delim = ";", show_col_types = FALSE)
cat("rows:", nrow(raw), "  baskets:", n_distinct(raw$transaction_id), "  items:", n_distinct(raw$article_name), "\n")
```

## R — `arules`

### Build the transaction object

```{r}
#| label: build-transactions
library(arules)

baskets <- split(raw$article_name, raw$transaction_id)
baskets <- lapply(baskets, unique)  # ensure no within-basket duplicates
trans   <- as(baskets, "transactions")
summary(trans)
```

### Item frequency

The 15 most-purchased products:

```{r}
#| label: fig-item-frequency
#| fig-cap: "Top 15 articles by basket appearance"
itemFrequencyPlot(trans, topN = 15, type = "absolute",
                  col = "steelblue", main = "")
```

`dining_chair` leads — exactly as designed: chairs are bought in sets of 2/4/6 and they ride along on every dining-table purchase.

### Mining rules

```{r}
#| label: apriori-r
rules <- apriori(trans,
                 parameter = list(support = 0.001, confidence = 0.5),
                 control   = list(verbose = FALSE))
cat("Total rules found:", length(rules), "\n")
```

Apriori returns hundreds of rules at this threshold. Sorting by confidence puts the most reliable ones first — but the very top is dominated by *complex* rules of the form `{A, B, C, D} ⇒ E`. These are technically high-confidence (often 100%) but cover so few baskets that they're noisy.

The headline patterns are the **simple** rules: one item on the left, one on the right.

```{r}
#| label: simple-rules-r
simple <- subset(rules, size(rules) == 2)
simple_sorted <- sort(simple, by = "confidence", decreasing = TRUE)
inspect(head(simple_sorted, 10))
```

These recover the co-purchase patterns we encoded in the synthesis:

- `dining_table ⇒ dining_chair` at ~86% confidence
- `kitchen_table ⇒ kitchen_chair` at ~75%
- `bed ⇒ mattress` at ~69%
- `desk ⇒ office_chair` at ~72%
- `garden_table ⇒ garden_chair` at ~70%

Lift values in the 4–10× range mean these pairs co-occur several times more often than chance.

### What about the complex rules?

Multi-item rules can hit 100% confidence because they're highly specific — `{bed, mattress, table_extension} ⇒ dining_table` triggers only on the rare baskets where someone happens to buy bedroom and dining furniture together. They're not noise, but they're more interesting as patterns to investigate than as rules to act on.

### Visualizing rules

```{r}
#| label: fig-rules-scatter
#| fig-cap: "Rule landscape: each point is one rule. Top-right corner = high support **and** high confidence (the most actionable rules)."
library(arulesViz)
plot(rules, method = "scatterplot",
     measure = c("support", "confidence"), shading = "lift",
     engine = "ggplot2")
```

## Python — `mlxtend`

The same analysis with the standard Python stack:

```{python}
#| label: apriori-python
import pandas as pd
from mlxtend.preprocessing import TransactionEncoder
from mlxtend.frequent_patterns import apriori, association_rules

df = pd.read_csv("data/synthetic/transactions.csv", sep=";")
baskets = df.groupby("transaction_id")["article_name"].apply(lambda s: list(set(s))).tolist()

te = TransactionEncoder()
basket_matrix = pd.DataFrame(te.fit_transform(baskets), columns=te.columns_)

freq_items = apriori(basket_matrix, min_support=0.001, use_colnames=True)
rules = association_rules(
    freq_items, num_itemsets=len(basket_matrix),
    metric="confidence", min_threshold=0.5,
)
print(f"Total rules found: {len(rules)}")
```

Same trick as in R — show the simple rules (one item on each side) first:

```{python}
#| label: simple-rules-python
is_simple = (rules["antecedents"].apply(len) == 1) & (rules["consequents"].apply(len) == 1)
simple = (
    rules.loc[is_simple]
    .sort_values("confidence", ascending=False)
    .assign(
        antecedent=lambda d: d["antecedents"].apply(lambda s: next(iter(s))),
        consequent=lambda d: d["consequents"].apply(lambda s: next(iter(s))),
    )
    [["antecedent", "consequent", "support", "confidence", "lift"]]
    .reset_index(drop=True)
)
simple.head(10).round({"support": 4, "confidence": 3, "lift": 2})
```

```{python}
#| label: fig-rules-python
#| fig-cap: "Same scatter view in Python: support × confidence, colored by lift."
import matplotlib.pyplot as plt
import seaborn as sns

fig, ax = plt.subplots(figsize=(8, 5))
sns.scatterplot(data=rules, x="support", y="confidence", hue="lift",
                palette="viridis", ax=ax)
ax.set_xscale("log")
ax.set_xlabel("Support (log scale)")
ax.set_ylabel("Confidence")
plt.tight_layout()
plt.show()
```

## R vs. Python — same answers?

Both implementations find the same total rule count and recover the same simple-rule headline patterns at identical support / confidence values — the algorithm is deterministic, not the implementation. The exact ordering of complex rules can differ when many tie on confidence (1.0), but that's a tiebreaker artifact, not a difference in what the algorithms find.

### Implementation differences worth knowing

| | `arules` (R) | `mlxtend` (Python) |
|---|---|---|
| Internal representation | sparse C-level transaction matrix | one-hot DataFrame |
| Performance on large data | very fast (10× ahead of mlxtend on 100k+ baskets) | acceptable up to ~10k baskets |
| Visualization | rich (`arulesViz`: scatter, graph, parallel coordinates, matrix) | DIY with matplotlib / networkx |
| Subsetting / pruning rules | first-class (`subset(rules, ...)`) | pandas filtering on the rules frame |
| Ecosystem fit | preferred when downstream is R / shiny | preferred when downstream is FastAPI / serving |

For exploratory work and visualization the R toolkit wins; for embedding into a Python service `mlxtend` is the easier deploy.

## Takeaway

The Apriori algorithm recovers the engineered co-purchase patterns cleanly in both languages. For a real retailer these rules drive concrete decisions: *which products to display together, what to bundle, what to suggest as upsell at checkout.* Practically: `dining_table` shoppers should never leave the store without seeing chairs, and the bed display should sit next to the mattress and nightstand sections.
