---
title: "RFM Clustering — Recency × Frequency × Monetary"
---

[RFM analysis](https://en.wikipedia.org/wiki/RFM_(market_research)) is a 30-year-old segmentation framework, originally for direct marketing. Each subject (customer or, here, *product*) is summarized by three numbers:

- **Recency** — how long ago was it last purchased? (lower = better)
- **Frequency** — how often was it purchased? (higher = better)
- **Monetary** — how much revenue did it generate? (higher = better)

The classic application is customer scoring; we apply the same logic at the article level — *which products in the catalog are healthy, which are dying, which need attention?*

```{r}
#| label: setup
#| message: false
library(readr)
library(dplyr)
library(ggplot2)
library(plotly)
library(rpart)
library(rpart.plot)

set.seed(42)
```

## Compute R, F, M per article

Reference date is the last day of the data window. Recency is measured in months since the article's last sale.

```{r}
#| label: rfm-features
.data_path <- if (file.exists("data/raw/transactions.csv")) "data/raw/transactions.csv" else "data/synthetic/transactions.csv"
raw <- read_delim(.data_path, delim = ";", show_col_types = FALSE)

REF_DATE <- max(raw$date)

rfm <- raw |>
  group_by(article_name) |>
  summarise(
    frequency = n(),
    value     = sum(gross_price),
    recency   = as.numeric(REF_DATE - max(date)) / 30.4375,  # months
    .groups   = "drop"
  )

cat(sprintf("Reference date: %s | %d articles, recency range [%.1f, %.1f] months, value range [%.0f, %.0f] EUR\n",
            format(REF_DATE), nrow(rfm), min(rfm$recency), max(rfm$recency),
            min(rfm$value), max(rfm$value)))
```

A quick look at the marginal distributions:

```{r}
#| label: fig-rfm-distributions
#| fig-cap: "Marginal distributions of recency (months since last sale), frequency (line-item count), and monetary value (EUR)."
#| fig-width: 11
#| fig-height: 3.2
library(patchwork)
p1 <- ggplot(rfm, aes(recency))   + geom_histogram(bins = 15, fill = "steelblue") + labs(x = "recency (months)")
p2 <- ggplot(rfm, aes(frequency)) + geom_histogram(bins = 15, fill = "darkorange") + labs(x = "frequency")
p3 <- ggplot(rfm, aes(value))     + geom_histogram(bins = 15, fill = "seagreen") + labs(x = "value (EUR)")
p1 + p2 + p3
```

Most articles have very recent last sales — the data window ends at the reference date and most products are still selling. A small tail of articles last sold months ago — on synthetic data these are the declining products engineered into the synthesis (`filing_cabinet`, `dvd_player`); on real data they are end-of-life SKUs.

## How many clusters?

```{r}
#| label: fig-elbow
#| fig-cap: "Elbow plot: the within-cluster sum of squares drops sharply through k=4, flattening after."
features <- scale(rfm[, c("frequency", "recency", "value")])

inertias <- sapply(1:10, function(k) {
  kmeans(features, centers = k, nstart = 10)$tot.withinss
})

ggplot(data.frame(k = 1:10, ss = inertias), aes(k, ss)) +
  geom_line() + geom_point(size = 2) +
  scale_x_continuous(breaks = 1:10) +
  labs(x = "Number of clusters", y = "Within-cluster SS")
```

The bend lands around **k = 4** — same as the BCG analysis, suggesting the catalog naturally falls into roughly four product archetypes.

## Fit and rank clusters

```{r}
#| label: fit-kmeans
K <- 4
fit <- kmeans(features, centers = K, nstart = 25)
rfm$cluster <- fit$cluster

# Rank clusters: higher value & frequency, lower recency = better
centers <- as.data.frame(fit$centers) |>
  tibble::rowid_to_column("cluster") |>
  mutate(weight = (frequency + value - recency) / 3) |>
  arrange(desc(weight)) |>
  mutate(rank = row_number())

rfm <- rfm |>
  left_join(centers |> select(cluster, rank), by = "cluster")

centers |> select(cluster, rank, recency, frequency, value, weight) |> round(3)
```

## The 3D view

Plotly gives us a rotatable, zoomable 3D scatter — much more useful than a static projection when three axes matter.

```{r}
#| label: fig-3d-rfm
#| fig-cap: "Interactive 3D RFM space. Drag to rotate, scroll to zoom, hover for product names. Cluster numbers are ranked: 1 = top performers (frequent, valuable, recent)."
plot_ly(
  rfm,
  x = ~recency, y = ~frequency, z = ~value,
  color = ~factor(rank),
  colors = c("#27ae60", "#3498db", "#f39c12", "#c0392b"),
  text = ~article_name,
  hovertemplate = paste0(
    "<b>%{text}</b><br>",
    "recency: %{x:.1f} mo<br>",
    "frequency: %{y}<br>",
    "value: €%{z:,.0f}<extra></extra>"
  ),
  type = "scatter3d", mode = "markers",
  marker = list(size = 6, opacity = 0.85)
) |>
  layout(
    scene = list(
      xaxis = list(title = "Recency (months)"),
      yaxis = list(title = "Frequency"),
      zaxis = list(title = "Monetary value (EUR)")
    ),
    legend = list(title = list(text = "<b>Rank</b>"))
  )
```

## Cluster narrative

```{r}
#| label: cluster-summary
rfm |>
  group_by(rank) |>
  summarise(
    n_products = n(),
    mean_recency   = round(mean(recency),   2),
    mean_frequency = round(mean(frequency), 0),
    mean_value     = round(mean(value),     0),
    examples = paste(head(article_name[order(-value)], 4), collapse = ", "),
    .groups = "drop"
  )
```

Reading the table:

- **Rank 1** — recent, frequently sold, high revenue. The flagship products (sofas, beds, dining tables and their high-frequency partners). Marketing should protect these.
- **Rank 2 / 3** — middle of the pack. Either decent volume but lower revenue (smaller-ticket items like chairs and decor), or solid revenue but lower volume.
- **Rank 4** — high recency (i.e. *not* recent — last sold months ago), low frequency, low value. The end-of-life products. Candidates for delisting or replacement.

## A simple rule from the clusters

The decision tree reduces the 4-cluster structure to a few interpretable splits:

```{r}
#| label: fig-tree
#| fig-cap: "Decision tree predicting cluster rank from raw R/F/M values. The thresholds are the boundaries the k-means actually picked, made readable."
tree <- rpart(rank ~ recency + frequency + value, data = rfm,
              method = "class", control = rpart.control(maxdepth = 4))
rpart.plot(tree, type = 2, extra = 100, fallen.leaves = TRUE,
           box.palette = "BuOr", main = "")
```

Codified as a few `if`-thresholds, this is a deployable scoring rule that needs no clustering library at runtime.

## Coarse view — RFM at Department level

The Family-level scatter is detailed enough for catalog operations. For executive reporting the more digestible aggregation is **Department** — six store-floor sections instead of dozens of products. With six points clustering becomes pointless; we just compute R/F/M per department and read the result directly.

```{r}
#| label: rfm-dept
raw_dept <- raw |>
  filter(!is.na(department), department != "") |>
  group_by(department) |>
  summarise(
    frequency = n(),
    value     = sum(gross_price),
    recency   = as.numeric(REF_DATE - max(date)) / 30.4375,
    .groups   = "drop"
  ) |>
  arrange(desc(value))

raw_dept
```

```{r}
#| label: fig-rfm-dept
#| fig-cap: "Department-level RFM. Bubble size encodes frequency; positions encode recency × monetary value. Useful for budget conversations."
ggplot(raw_dept, aes(recency, value, size = frequency, label = department)) +
  geom_point(color = "#3498db", alpha = 0.7) +
  geom_text(size = 4, fontface = "bold", vjust = -1.4) +
  scale_size(range = c(8, 20), guide = "none") +
  labs(x = "Recency (months since last sale)", y = "Total revenue (EUR)") +
  theme_minimal()
```

Department-level readings are coarser by design — you lose the per-product detail but gain a cleaner narrative for stakeholders who don't want to parse a 40-point 3D scatter.

## Takeaway

The k=4 RFM segmentation cleanly separates the catalog into a healthy core, a long middle, and a tail of dying products. The framework was originally designed for customers but transfers directly to articles — and the same techniques work for *any* entity you can summarize as (recency, frequency, monetary).

For the standalone Python version of this same analysis with a more interactive 3D experience, see [`notebooks/rfm_clustering.ipynb`](notebooks/rfm_clustering.ipynb).
